---
title: "Biodiversity analysis"
author: "Yining Sun"
date: "2023-04-13"
output: html_document
editor_options: 
  chunk_output_type: console
---

# load packages
```{r}
# install.packages("iNEXT")

# install.packages("ggpubr") this package has function that can do statistical tests in ggplot

# efficiently load packages
pacman::p_load(phyloseq, iNEXT, ggpubr, tidyverse, install = FALSE)

# load functions and color preferences
source("./code/functions.R")
source("./code/colors_and_shapes.R")
```


# load data
```{r}
load("./data/preprocessed_physeq.RData")

preprocessed_physeq

```


# run biodiversity analysis
## work with the iNEXT packages
```{r iNEXT-div-estimation}
# prepare data for iNEXT
iNEXT_input_df <-
  preprocessed_physeq %>%
  otu_table() %>%
  data.frame()


# run iNEXT on the data
# warning: ~ 5 mins to run
# calculate diversity with iNEXT
set.seed(0402)


# iNEXT_data <- iNEXT(iNEXT_input_df, q = c(0, 1, 2), datatype = "abundance")
# save(iNEXT_data, file = "./data/iNEXT_MissionAransas.RData")

# q = 0, 1, 2
# q = 0 -> total number of taxa
# q = 1 -> number of common taxa, equal to shannon diversity
# q = 2 -> number of dominant taxa


load("./data/iNEXT_MissionAransas.RData")
# note: the obkect is called "iNEXT_data"
str(iNEXT_data)
typeof(iNEXT_data)


# make workable df
div_iNEXT_df <-
  iNEXT_data$AsyEst %>%
  dplyr::rename(names = Assemblage) %>%
  make_MA2_metadata()
```


# plot diversity in samples
```{r}
fraction_colors

# Set colors for manual plotting with iNEXT
background_col_FL <- "goldenrod1"
background_col_PA <- "firebrick3"
background_col_WH <- "darkorange2"

###
# Prepare data to color the figure for iNEXT
dat <- colnames(iNEXT_input_df) %>%    
  data.frame() 

colnames(dat)[1] <- "names"     

sub_metadata <- dat %>% make_MA2_metadata()

# Add the colors for the plotting!
dat_iNEXT <- dat %>%      
  left_join(sub_metadata, by = "names") %>%    
  mutate(fraction_color = ifelse(fraction == "Particle", background_col_PA,                                  ifelse(fraction == "Free", background_col_FL,                                        background_col_WH)))

# Now plot it all :) 
# Plot the rarefaction/extrapolation curve (type = 1)
rarefaction_fixed_plot <-   ggiNEXT(iNEXT_data, type=1, facet.var="Order.q") +   
  facet_wrap(~Order.q, scales="fixed") +  
  scale_color_manual(values = dat_iNEXT$fraction_color,  guide = FALSE) +  
  scale_fill_manual(values = dat_iNEXT$fraction_color, guide = FALSE) +  
  theme(legend.position = "none") +   
  labs(x = "Number of Sequences")

rarefaction_fixed_plot

# Save the file using ggsave()! 
ggsave(rarefaction_fixed_plot,        
       filename = "figures/rarefaction_fixed.png",       
       units = "in", height = 3.5, width = 8)
```



# pull out specific samples and plot it
```{r}
# explore the iNEXT object
str(iNEXT_data)

# pull out information on rarefaction
str(iNEXT_data$iNextEst$size_based)
iNEXT_data[[2]][1]


# look into the observed diversity values and check them
obs_rich_df <-
  iNEXT_data$iNextEst$size_based %>%
  # extract the richness from the data
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Observed") %>%
  dplyr::rename(names = Assemblage) %>%
  # rename the X's from the sample names
  mutate(names = gsub("X", "", names)) %>%
  make_MA_metadata()
 

# manually plot the rarefaction
iNEXT_data$iNextEst$size_based %>%
  # extract the richness from the data
  dplyr::filter(Order.q == 0) %>%
  dplyr::filter(Method == "Rarefaction") %>%
  dplyr::rename(names = Assemblage) %>%
  # rename the X's from the sample names
  mutate(names = gsub("X", "", names)) %>%
  make_MA_metadata() %>%
  ggplot(aes(x = m, y = qD, color = fraction, group = names)) + 
  geom_line() + 
  geom_point(data = obs_rich_df, aes(x = m, y = qD, color = fraction)) +
  facet_grid(.~fraction) +
  scale_color_manual(values = fraction_colors)


## the sequences needed to reach the flatness is called sequence depth, the smaller the sequence depth is, the fewer sequence the sample needed to flat


```
Sample 20210615.MA.ABB2F has a much lower richness compared to all the other smaples, will keep an eye out for this smaple.


# completness curve
```{r}

ggiNEXT(iNEXT_data, type = 2) + 
  scale_color_manual(values = dat_iNEXT$fraction_color,  guide = FALSE) +  
  scale_fill_manual(values = dat_iNEXT$fraction_color, guide = FALSE) +  
  # zoom in
  scale_x_continuous(limit = c(0, 2000)) +
  theme(legend.position = "none") 


```


# boxplot of diversity
```{r boxplot-div}

iNEXT_data$AsyEst %>%
  head()

obs_rich_df %>%
  dplyr::filter(fraction == "Whole") %>%
  ggplot(aes(x = station, y = qD, color = station)) +
  geom_jitter(size = 2.5) +
  geom_boxplot(aes(fill = station), alpha = 0.5, outlier.shape = NA) +
  scale_fill_manual(values = station_colors) +
  scale_color_manual(values = station_colors) +
  theme(axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1)) +
  stat_kruskal_test()

# Kruskal test
kruskal.test(qD ~ station, 
             data = dplyr::filter(obs_rich_df, fraction == "Whole"))
```

# tests for sigs
## multiple comparison:
### Kruskal-walls (non-parametric, does not require normal distribution)
### ANOVA (parametric, require normal distribution)
## two paired samples
### Wilcoxon (non-parametric)
### t test (parametric)



