---
title: "Phyloseq Pre-Processing"
author: "Yining Sun"
date: "2023-03-28"
output: html_document
---

# Goals of our Pre-Processing File

We will work here tin this document to remove any unnecessay ASVs or samples, such as:

1. Mitochondria ASVs
2. Chloroplast ASVs
3a. ASVs found in the controls
3b Controls
4. Mock ASVs/Mock Community


## Load packages and Functions
```{r load packages-functions}
# set wd
setwd("/local/workdir/ys636/BIOMI6300_Amplicon_Analysis")

# load packages
pacman::p_load(tidyverse, phyloseq, install = FALSE)

# source functions
source("code/functions.R")

```


## Load the data
```{r load-data}

# load in the raw_physeq data from DADA2 Practice.Rmd
load("/local/workdir/ys636/BIOMI6300_Amplicon_Analysis/data/raw_physeq.RData")

# Take a look at the S4 object
raw_physeq

# Tell us about your phyloseq
str(raw_physeq)
typeof(raw_physeq)

View(raw_physeq@otu_table)
View(raw_physeq@tax_table)
View(data.frame(sample_data(raw_physeq)))

```


# Remove unnecessary taxa
```{r rm-ASVs}

# 1. Mitochondria ASVs
# 2. Chloroplast ASVs
# 3a. ASVs found in the controls
# 3b Controls
# 4. Mock ASVs/Mock Community

# test_physeq <-
  # raw_physeq %>% 
   ## remove the mitochondria ASVs
  # subset_taxa(Family != "Mitochondria") %>%
   ## remove the chloroplast ASVs
  # subset_taxa(Order != "Chloroplast")


raw_physeq %>%
  subset_taxa(Family != "Mitochondria") %>%
  subset_taxa(Order != "Chloroplast") %>%
  prune_samples((sample_names(.)) %!in% c("CJ-V08-P", "MockZyomPos"), .) %>%
  # now we will remove any ASVs of count 0
  prune_taxa(taxa_sums(.) > 0, .)
  

noMitoChloroCJMock_physeq

# how many taxa have we removed so far?
num_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noMitoChloroCJMock_physeq)
# proportion
prop_ASVs_rm <- ntaxa(raw_physeq)/ntaxa(noMitoChloroCJMock_physeq)

```


Methods:
 
Mitochondria, chloroplasts, and mock community were removed resulting in the loss of `r 1-prop_ASV_rm`% of the data


## Remove control samples
```{r}
# 3b. Control samples

# create a vector with the control samples
control_samples <- c("WaterControl", "022um_Control", "3um-Control", "DNA-Ext-Control")

# what taxa are in the control samples?
noMitoChloroCJMock_physeq %>%
  # identify control samples
  subset_samples(., (names %!in% control_samples)) %>%
  # make sure to remove the ASVs that were not in the controls
  prune_taxa(taxa_sums(.) > 0, .)

control_physeq %>%
  plot_bar(., "ASV", fill - "ASV")



# control ASVs
control_ASVs <- data.frame(control_physeq@tax_table)$ASV

View(control_physeq@tax_table)


```



