---
title: "Phyloseq Pre-Processing"
author: "Yining Sun"
date: "2023-03-28"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Goals of our Pre-Processing File

We will work here tin this document to remove any unnecessay ASVs or samples, such as:

1. Mitochondria ASVs
2. Chloroplast ASVs
3a. ASVs found in the controls
3b Controls
4. Mock ASVs/Mock Community
5. samples that have too low # of reads for reasonable ecological analysis 


## Load packages and Functions
```{r load packages-functions}
# set wd
setwd("/local/workdir/ys636/BIOMI6300_Amplicon_Analysis")

# load packages
pacman::p_load(tidyverse, phyloseq, install = FALSE)

# source functions
source("code/functions.R")
source("code/colors_and_shapes.R")
```

## Load the data
```{r load-data}

# load in the raw_physeq data from DADA2 Practice.Rmd
load("./data/raw_physeq.RData")

# Take a look at the S4 object
raw_phyloseq

# Tell us about your phyloseq
str(raw_phyloseq)
typeof(raw_phyloseq)

View(raw_phyloseq@otu_table)
View(raw_phyloseq@tax_table)
View(data.frame(sample_data(raw_phyloseq)))

```


# Remove unnecessary taxa
```{r rm-ASVs}

# 1. Mitochondria ASVs
# 2. Chloroplast ASVs
# 3a. ASVs found in the controls
# 3b Controls
# 4. Mock ASVs/Mock Community

# test_physeq <-
  # raw_physeq %>% 
   ## remove the mitochondria ASVs
  # subset_taxa(Family != "Mitochondria") %>%
   ## remove the chloroplast ASVs
  # subset_taxa(Order != "Chloroplast")

noMitoChloroCJMock_physeq <- 
  raw_phyloseq %>%
  subset_taxa(Family != "Mitochondria" | is.na(Family)) %>%
  subset_taxa(Order != "Chloroplast" | is.na(Order)) %>%
  prune_samples((sample_names(.)) %!in% c("CJ-V08-P", "MockZyomPos"), .) %>%
  # now we will remove any ASVs of count 0
  prune_taxa(taxa_sums(.) > 0, .)
  

noMitoChloroCJMock_physeq

# how many taxa have we removed so far?
num_ASVs_rm <- ntaxa(raw_phyloseq) - ntaxa(noMitoChloroCJMock_physeq)
# proportion
prop_ASVs_rm <- ntaxa(raw_phyloseq)/ntaxa(noMitoChloroCJMock_physeq)

```


Methods:
 
Mitochondria, chloroplasts, and mock community were removed resulting in the loss of `r 1-prop_ASV_rm`% of the data


## Remove control samples
```{r}
# 3b. Control samples

# create a vector with the control samples
control_samples <- c("WaterControl", "022um_Control", "3um-Control", "DNA-Ext-Control")

# what taxa are in the control samples?
control_phyloseq <-
  noMitoChloroCJMock_physeq %>%
  # identify control samples
  subset_samples(., (names %in% control_samples)) %>%
  # make sure to remove the ASVs that were not in the controls
  prune_taxa(taxa_sums(.) > 0, .)

control_phyloseq %>%
  plot_bar(., "ASV", fill = "ASV")



# control ASVs
control_ASVs <- data.frame(control_phyloseq@tax_table)$ASV

View(control_phyloseq@tax_table)





```


# Dealing with Controls
```{r control_asv}
## REMOVE OTUs that were in the controls 
control_samples <- c("WaterControl", "DNA-Ext-Control", "3um-Control", "022um-Control")

#################################
# Which taxa are in the control samples? 
control_physeq <- 
  noMitoChloroCJMock_physeq %>%
  subset_samples(., (names %in% control_samples)) %>%
  # Now remove ASVs that were not in control samples 
  prune_taxa(taxa_sums(.) > 0, .) 

# Which ASVs are in the control? 
control_ASVs <- data.frame(control_physeq@tax_table)$ASV

################################# ASV ANALYSIS YOU CAN SKIP #################################
controlASVs_physeq <- 
  noMitoChloroCJMock_physeq %>%
  subset_taxa(., ASV %in% control_ASVs)

# Let's take a look at the plot 
controlASVs_physeq %>%
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# From this plot, it looks like ASV_1 is REAL and perhaps ASV_20? Let's check 
controlASVs_physeq %>%
  subset_taxa(., ASV %in% c("ASV_3", "ASV_1", "ASV_20")) %>% # 
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  labs(title = "Should we keep these ASVs?") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))

# All the other OTUs 
controlASVs_physeq %>%
  subset_taxa(., ASV %!in% c("ASV_3", "ASV_1", "ASV_20")) %>% # ALSO SHOW "ASV_20"
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
################################################################################

# FROM ABOVE ANALYSIS 
# "ASV_1" and "ASV_20" MUST BE KEPT IN THE DATA! 
# Make a vector of all ASVs to remove
control_ASVs_toREMOVE <- dplyr::setdiff(control_ASVs, c("ASV_1", "ASV_20"))

# Now make a new phyloseq object without these! 
length(control_ASVs_toREMOVE)
length(control_ASVs)

# Create Phyloseq object that has control samples removed. 
noCONTROLSMitoChloroCJMock_physeq <- 
  noMitoChloroCJMock_physeq %>%
  # Now remove ASVs found in control samples samples, EXCEPT ASV_1 and ASV_20
  subset_taxa(., (ASV %!in% control_ASVs_toREMOVE)) %>%
  # Now remove control samples
  subset_samples(., (names %!in% control_samples)) %>%
  # Remove ASVs with 0 counts. 
  prune_taxa(taxa_sums(.) > 0, .) 

# After filtering we have removed a total of how many ASVs? 
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noCONTROLSMitoChloroCJMock_physeq)
total_ASVs_rm
```


# Dealing with Controls
```{r control_asv}
## REMOVE OTUs that were in the controls 
control_samples <- c("WaterControl", "DNA-Ext-Control", "3um-Control", "022um-Control")

#################################
# Which taxa are in the control samples? 
control_physeq <- 
  noMitoChloroCJMock_physeq %>%
  subset_samples(., (names %in% control_samples)) %>%
  # Now remove ASVs that were not in control samples 
  prune_taxa(taxa_sums(.) > 0, .) 

# Which ASVs are in the control? 
control_ASVs <- data.frame(control_physeq@tax_table)$ASV

# which ASVs do we want to keep in our datasets
################################# ASV ANALYSIS in the controls #################################
controlASVs_physeq <- 
  noMitoChloroCJMock_physeq %>%
  subset_taxa(., ASV %in% control_ASVs)

# Let's take a look at the plot 
controlASVs_physeq %>%
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

# From this plot, it looks like ASV_1 is REAL and perhaps ASV_20? Let's check 
controlASVs_physeq %>%
  subset_taxa(., ASV %in% c("ASV_3", "ASV_1", "ASV_20")) %>% # 
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  labs(title = "Should we keep these ASVs?") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))

# All the other OTUs 
controlASVs_physeq %>%
  subset_taxa(., ASV %!in% c("ASV_3", "ASV_1", "ASV_20")) %>% # ALSO SHOW "ASV_20"
  plot_bar(., "ASV", fill="ASV", facet_grid=Sample_or_Control~.) + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
################################################################################

# FROM ABOVE ANALYSIS 
# "ASV_1" and "ASV_20" MUST BE KEPT IN THE DATA! 
# Make a vector of all ASVs to remove
control_ASVs_toREMOVE <- dplyr::setdiff(control_ASVs, c("ASV_1", "ASV_20"))

# Now make a new phyloseq object without these! 
length(control_ASVs_toREMOVE)
length(control_ASVs)

# Create Phyloseq object that has control samples removed. 
noCONTROLSMitoChloroCJMock_physeq <- 
  noMitoChloroCJMock_physeq %>%
  # Now remove ASVs found in control samples samples, EXCEPT ASV_1 and ASV_20
  subset_taxa(., (ASV %!in% control_ASVs_toREMOVE)) %>%
  # Now remove control samples
  subset_samples(., (names %!in% control_samples)) %>%
  # Remove ASVs with 0 counts. 
  prune_taxa(taxa_sums(.) > 0, .) 

# After filtering we have removed a total of how many ASVs? 
total_ASVs_rm <- ntaxa(raw_physeq) - ntaxa(noCONTROLSMitoChloroCJMock_physeq)
total_ASVs_rm
```


## view ASVs in both controls and samples, if one ASV is high in the control while low in the samples, it should be removed.

### !filter=0.25% | samples with low number of reads



# evaluate sequence depth
```{r}
# look at the phyloseq object
noCONTROLSMitoChloroCJMock_physeq

# calculate the sum of the seqs in each sample
seqSums_df <- 
  noCONTROLSMitoChloroCJMock_physeq %>%
  otu_table() %>%
  # take the sum of each column of the ASV matrix
  colSums() %>%
  data.frame() %>%
  rownames_to_column(var = "names") %>%
  make_MA_metadata()

head(seqSums_df)


# fix column name in 2nd column
colnames(seqSums_df)[2] <- "Total seqs"

head(seqSums_df)

# plot it
# histogram plot
seqSums_df %>%
  ggplot(aes(x = TotalSeqs)) +
  geom_histogram(bins = 50) +
  labs(title = "Histogram of raw seq counts per sample")


# density plot
seqSums_df %>%
  ggplot(aes(x = TotalSeqs, fill = fraction)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = fraction_colors)


```



# remove sample with too few reads
```{r}
preprocessed_physeq <-
  noCONTROLSMitoChloroCJMock_physeq %>%
  subset_samples(., names != "20210615-MA-SCB2F")

min(sample_sums(preprocessed_physeq))
```


# save preprocessed phyloseq
```{r}
save(preprocessed_physeq, file = "./data/preprocessed_physeq.RData")
```


